---
title: "Data-exploration.qmd"
format: html
editor: visual
---

## Data Exploration on Bears

setting things up

```{r}
rm(list=ls())
library(here)
library(tidyverse)
library(ggfortify)
```

```{r}
#read data in 
bears <- read_csv("data/Short_et_al_Data1 (1).csv")
```

```{r}
glimpse(bears)
summary(bears)

```

```{r}
#change character variables to factor variables 
bears$Exposure <- as.factor(bears$Exposure)
bears$PeoplePresent <- as.factor(bears$PeoplePresent)
bears$Treatment <- as.factor(bears$Treatment)
```

```{r}
#plot the relationship between the detection of bears and the salmon biomass to see the correlation 
ggplot(bears, aes(x= Detections, y= SalmonBiomass )) + geom_point(size = 5) +  theme_bw()
```

From this, there doesn't seem to be any huge outliers due to mistakes in the data. I will not be removing any points and keeping the data as is. You can also detect from this plot a positive relationship between salmon biomass and the amount of bears.

```{r}
#making a histogram for comparing years to bear 
ggplot(bears, aes(x= Year)) + geom_histogram(binwidth = .25) 
```

From the data you can expect to see the relationship between the year it is and the bear count.

```{r}
#finding the mean and standard deviation 
salmon <- bears %>% group_by(Detections) %>% summarise(
  mean_s = mean(SalmonBiomass, na.rm= T), 
  sd_s= sd(SalmonBiomass, na.rm= T)
)
View(salmon)
```

```{r}
count <- bears %>% group_by(Year) %>% summarise(
  mean_c = mean(Detections, na.rm= T), 
  sd_c= sd(Detections, na.rm= T)
)
View(count)
```
